Think deep for reef resilience

### 2025-07-26 paper/paper_draft.qmd
Proof reading edits for discussion section.

The large differences in cover timeseries seen across the GCMs suggest that it is important
to consider this uncertainty in future conditions when assessing possible biological
responses to change. This is supported by recent literature suggesting an effective approach
to capturing climate change uncertainty is to model climate change responses using a range
of climate models separately [@shepherd2018]. Despite large differences in coral cover
across GCMs, the patterns identified between low, medium and high clusters of reefs suggest
that the importance of depth refugia is consistent across scenarios.

Reefs in different clusters largely do not differ in mean modelled DHW levels experienced,
and exhibit overall similar DHW timeseries. While these reefs experience similar magnitude
DHWs based on SST, the proportion of the reefs' coral population impacted by heat stress
events depends on reef depth. Our results suggest that within shallow-water reef ecosystems
(non-mesophotic, depth \< 30m), depth contributes greatly to reef resilience and the ability
of the reef to maintain positive carbonate budgets under future conditions. While similar
findings have been demonstrated in the eastern Pacific [@smith2014], depth refugia analyses
have often focussed on the potential of mesophotic reefs (\> 30m deep) as thermal refugia
[@bongaerts2010; @holstein2015; @semmler2017]. In contrast, Venegas et al. [-@venegas2019]
found no association between depth and the alleviation of heat stress across a range of
shallow reefs in the Pacific. Reefs at shallow depths experience no, or few, years with
coral cover exceeding levels likely required to maintain positive carbonate budgets.
Movement of shallow reefs towards negative cabonate budgets under increasing future thermal
stress over time may contribute to phase shifts in reef dominant CaCO~3~ production.
Shifting from coral driven to coralline encrusting algae driven production, or to bioerosion
dominated systems with rapid erosion of physical reef frameworks [@dudgeon2010; @perry2008].
While reefs at increased depths may be able to maintain coral driven CaCO~3~ production and
accumulation into the future for longer.

Overall, these findings support the hypothesis that reefs at greater depths will experience
refugia benefits and more prolonged resilience under climate change compared to shallower
reefs. Additionally, [@liu2024] suggest that rubble instability is inversely correlated with
reef depth, possibly limiting the juvenile recruitment capacity of shallow reefs. A caveat
to these findings is the quality of the bathymetry data used to calculate median reef depth.
This data (GBRMPA [-@gbr10gb2020]) is derived from satellite imagery and is limited in its
alignment with the established GBRMPA reef outlines used in the investigation to
characterise reef locations. As a result, 842 reefs (22% of 3806 reefs) were removed from
the analyses before depth comparisons due to having less than 5% coverage of depth data.
Additionally, deriving estimates of depth from satellite imagery is prone to high degrees of
uncertainty. Data adopted for the given assessment are said to be accurate up to 15m from
Mean Sea Level [@gbr10gb2020]. Furthermore, an estimated median depth is determined for each
reef, using the established GBRMPA reef outlines to provide an indication of “true” reef
extent. Therefore, the estimates of reef depths are taken as a useful indication rather than
an accurate and precise report of reef depths across the GBR.

@@

Overall, our findings do not suggest that reefs in ‘high’ cover clusters have greater
thermal tolerance than reefs in lower coral cover clusters. This is in contrast with
literature suggesting corals have the capacity to adapt to increasing climate change
pressures [@humanes2022; @lachs2024], including the investigation by Bairos-Novak et al.
[-@bairos-novak2021] that revealed high heritability of bleaching and survival response
traits, used to inform the distributions of thermal tolerance traits used in this
investigation. The lack of influence of thermal tolerance on reef outcomes is likely due to
reefs being impacted by high levels of thermal stress early in timeseries and being unable
to build thermal tolerance to DHW levels greater than 9°C.

It is a common conclusion in coral reef ecology that a primary facet of reef resilience is
the degree of larval connectivity between reefs and the supply of recruits that a reef
receives following disturbance [@hock2017; @mumby2008; @torres2024]. In contrast, our
results suggest that there are few differences between resilient and non-resilient reefs in
connectivity metrics such as the ratio of outgoing to incoming larval connectivity and total
connectivity strength. Furthermore, our results suggest no relationship between connectivity
and the ability of reefs to maintain positive carbonate budgets under future conditions.
There has been contrasting evidence [@bode2018; @hock2017] surrounding whether recent
generation connectivity modelling is able to identify reefs that are important for recovery
processes, with findings dependent on the larval connectivity model used. These
investigations were performed using prior generations of larval connectivity models
parameterised for 2008-2013 [@hock2017] and 1996-2002 [@bode2018] time periods. Using
updated 1km resolution larval connectivity across the whole GBR, with a limited temporal
sample of particle flows, our results suggest that larval connectivity alone is not as
important for reef resilience and recovery as environmental refugia from thermal stress such
as depth.

Our findings on the importance of connectivity may differ due to the highly uncertain nature
of larval connectivity modelling [@vogt-vincent2023], the way larval flows and juvenile
recruitment are implemented in ADRIA - CoralBlox, and the magnitude of early bleaching
events decreasing the breeding capacity of strongly connected reefs. Analysing daily
connectivity data Vogt-Vincent et al. [-@vogt-vincent2023] suggest that time-mean
connectivity may not be suitable for capturing the seasonal - ecological time scale,
connectivity between locations due to the high daily variability in connectivity data. This
is supported by the connectivity data used in our investigation, which have large relative
standard deviations for connectivity between reefs over the 24 days used [@sfig-connectivity-rsd]. Using
alternative summarisation methods that capture ecological-timescale patterns in
coral larval dispersal, such as cumulated implicit connectivity [@vogt-vincent2023],
future investigations may find increased importance of both depth and reef connectivity for
resilience, in line with [@hock2017; @torres2024]. Further improvements and research in
connectivity modelling are necessary to provide a robust comparison of all factors
influencing reef resilience.

There are multiple caveats in how coral ecosystem dynamics are modelled in this
investigation. We model coral colonies as objects with circular diameter values and track
these diameters over timesteps through growth. Due to the availability of quality future
projected data we do not consider the impacts of waves, cyclones, currents and Crown of
Thorns Starfish (CoTS) on coral colony survival and bleaching [**Deaker et al**; **Massel and Done**]. Uncertainty in connectivity modelling outputs,
satellite-derived bathymetry data and future thermal stress projections are likely to
influence the results obtained in this investigation. Incorporating further improvements in
connectivity modelling and understanding of the influence of climate change on reef
connectivity would improve the comparison of connectivity as an important factor for reef
resilience. The DHW data used in this investigation have not been bias corrected for the
Great Barrier Reef region and are part of a wider Australian dataset [@dixon2022] developed
using statistical downscaling. We investigate a range of future climate responses, thermal
projections and ecological parameterisations of ADRIA - CoralBlox to account for many
possible futures, and our results suggest the factors that influence reef resilience are
consistent across these scenarios.

@@

Further analyses could build on the findings of this study by analysing the impacts of
prioritising reef depth in restoration intervention decision making. The ADRIA platform is
developed as a modelling tool to incorporate multi-criteria decision analysis (MCDA) and
ecosystem scenario modelling into reef restoration decision support. The platform allows for
the investigation of different reef characteristics to be targetted in decision making and
models the evolving outcomes as a result of reef interventions at various locations. The
capability of ADRIA-CoralBlox could be further expanded by incorporating the impacts of
waves, cyclones, CoTS and currents into reef resilience investigations [@lin2023;
@mcclanahan2005; @grimaldi2022; @liu2024]. Additionally, the investigation could be enhanced
by incorporating future ocean acidification into considerations of coral carbonate budgets
required for healthy habitat production.

@@

![Heatmaps indicating reef cluster assignments within each bioregion across the five GCMs.
Green indicates a low cover cluster assignemnt, orange represents a medium cover assignment
and blue represents assignment of a reef to a high coral cover cluster. The vertical width
of reef elements varies depending on the number of reefs within the bioregion (displayed in
the y axis). The x axis represents the different GCM levels (EC-Earth3-Veg, ACCESS-ESM1-5,
ACCESS-CM2, NorESM2-MM and
GFDL-CM4).](../outputs/ADRIA_results/HighResCoralStress/figs/GCM_bioregion_cluster_assignment.png){#fig-gcm-bioregion-cluster-assignment
width="`{julia} timeseries_width`"}

>>>>

The large differences in cover timeseries seen across the GCMs suggest that it is important
to consider this uncertainty in future conditions when assessing possible biological
responses to change. This is supported by recent literature suggesting an effective approach
to capturing climate change uncertainty is to model responses using a range of GCMs
separately [@shepherd2018]. Despite large differences in coral cover across GCMs, the
patterns identified between low, medium and high clusters of reefs suggest that the
importance of depth refugia is consistent across scenarios.

Reefs in different clusters largely do not differ in mean modelled DHW levels experienced,
and exhibit overall similar DHW timeseries. While these reefs experience similar magnitude
DHWs based on SST, the proportion of the reefs' coral population impacted by heat stress
events depends on reef depth. Our results suggest that within shallow-water reef ecosystems
(non-mesophotic, depth \< 30m), depth contributes greatly to reef resilience and the ability
of the reef to maintain positive carbonate budgets under future conditions. While similar
findings have been demonstrated in the eastern Pacific [@smith2014], depth refugia analyses
have often focussed on the potential of mesophotic reefs (\> 30m deep) as thermal refugia
[@bongaerts2010; @holstein2015; @semmler2017]. In contrast, Venegas et al. [-@venegas2019]
found no association between depth and the alleviation of heat stress across a range of
shallow reefs in the Pacific. Reefs at shallow depths experience no, or few, years with
coral cover exceeding levels likely required to maintain positive carbonate budgets.
Movement of shallow reefs towards negative cabonate budgets in the future may contribute to
phase shifts in reef dominant CaCO~3~ production. Shifting from coral driven to encrusting
algae driven production, or to bioerosion dominated systems with rapid erosion of physical
reef frameworks [@dudgeon2010; @perry2008]. While reefs at increased depths may be able to
maintain coral driven CaCO~3~ production and accumulation into the future for longer.

Overall, these findings support the hypothesis that reefs at greater depths will experience
refugia benefits and more prolonged resilience under climate change compared to shallower
reefs. Additionally, Lui, Callaghanand Baldock [-@liu2024] suggest that rubble instability
is inversely correlated with reef depth, possibly limiting the juvenile recruitment capacity
of shallow reefs. A caveat to these depth-resilience findings is the quality of the
bathymetry data used to calculate median reef depth. These data (GBRMPA [-@gbr10gb2020]) are
derived from satellite imagery and are limited in their alignment with the established
GBRMPA reef outlines used in the investigation to characterise reef locations. As a result,
842 reefs (22% of 3806 reefs) were removed from the depth-related analyses due to having
less than 5% coverage of depth data. Additionally, deriving estimates of depth from
satellite imagery is prone to high degrees of uncertainty. Data adopted for the given
assessment are said to be accurate up to 15m from Mean Sea Level [@gbr10gb2020].
Furthermore, an estimated median depth is determined for each reef, using the established
GBRMPA reef outlines to provide an indication of “true” reef extent. Therefore, the
estimates of reef depths are taken as a useful indication rather than an accurate and
precise report of reef depths across the GBR.

@@

Overall, our findings do not suggest that reefs in ‘high’ cover clusters have greater
thermal tolerance than reefs in lower coral cover clusters. This is in contrast with
literature suggesting corals have the capacity to adapt to increasing climate change
pressures [@humanes2022; @lachs2024], including the investigation by Bairos-Novak et al.
[-@bairos-novak2021] that revealed high heritability of bleaching and survival response
traits, used to inform thermal tolerance trait distributions in this investigation. The lack
of influence of thermal tolerance on reef outcomes is likely due to reefs being impacted by
high levels of thermal stress early in timeseries and being unable to build thermal
tolerance to DHW levels greater than 9°C-weeks.

It is a common conclusion in coral reef ecology that a primary facet of reef resilience is
the degree of larval connectivity between reefs and the supply of recruits that a reef
receives following disturbance [@hock2017; @mumby2008; @torres2024]. In contrast, our
results suggest that there are few differences between resilient and non-resilient reefs in
connectivity metrics such as the ratio of outgoing to incoming larval connectivity and total
connectivity strength. Furthermore, our results suggest no relationship between connectivity
and the ability of reefs to maintain positive carbonate budgets under future conditions.
There has been contrasting evidence [@bode2018; @hock2017] surrounding whether recent
generation connectivity modelling is able to identify reefs that are important for recovery
processes, with findings dependent on the larval connectivity model used. These
investigations were performed using prior generations of larval connectivity models
parameterised for 2008-2013 [@hock2017] and 1996-2002 [@bode2018] time periods. Using
updated 1km resolution larval connectivity across the whole GBR, with a limited temporal
sample of particle flows, our results suggest that larval connectivity alone is not as
important for reef resilience and recovery as environmental refugia such as depth.

Our findings on the importance of connectivity may differ due to the highly uncertain nature
of larval connectivity modelling [@vogt-vincent2023], the way larval flows and juvenile
recruitment are implemented in ADRIA - CoralBlox, and the magnitude of early bleaching
events decreasing the breeding capacity of strongly connected reefs. Analysing daily
connectivity data Vogt-Vincent et al. [-@vogt-vincent2023] suggest that time-mean
connectivity may not be suitable for capturing the seasonal - ecological time scale
connectivity between locations due to high daily variability in connectivity data. This is
supported by the connectivity data used in our investigation, which have large relative
standard deviations for connectivity between reefs over the 24 days used
[@sfig-connectivity-rsd]. Using alternative summarisation methods that capture
ecological-timescale patterns in coral larval dispersal, such as cumulated implicit
connectivity [@vogt-vincent2023], future investigations may find increased importance of
both depth and reef connectivity for resilience, in line with other investigations
[@hock2017; @torres2024]. Further improvements and research in connectivity modelling are
necessary to provide a robust comparison of all factors influencing reef resilience.

There are multiple caveats in how coral ecosystem dynamics are modelled in this
investigation. We model coral colonies as circular diameter values and track these diameters
over timesteps through growth. Due to the availability of quality future projected data we
do not consider the impacts of waves, cyclones, currents and Crown of Thorns Starfish (CoTS)
on coral colony survival and bleaching [@massel1993; @deaker2022]. Uncertainty in
connectivity modelling outputs, satellite-derived bathymetry data and future thermal stress
projections are likely to influence the results obtained in this investigation.
Incorporating further improvements in connectivity modelling and understanding of the
influence of climate change on reef connectivity would improve the comparison of
connectivity as an important factor for reef resilience. The DHW data used in this
investigation have not been bias corrected for the Great Barrier Reef region and are part of
a wider Australian dataset [@dixon2022] developed using statistical downscaling. We
investigate a range of future climate responses, thermal projections and ecological
parameterisations of ADRIA - CoralBlox to account for many possible futures, and our results
suggest the factors that influence reef resilience are consistent across these scenarios.

@@

Further analyses could build on the findings of this study by analysing the impacts of
prioritising reef depth in restoration intervention decision making. The ADRIA platform is
developed as a modelling tool to incorporate multi-criteria decision analysis (MCDA) and
ecosystem scenario modelling into reef restoration decision support. The platform allows for
the investigation of different reef characteristics to be targetted in decision making and
models the evolving outcomes as a result of interventions at various locations. The
capability of ADRIA-CoralBlox could be further expanded by incorporating the impacts of
waves, cyclones, CoTS and currents into future resilience investigations [@lin2023;
@mcclanahan2005; @grimaldi2022; @liu2024]. Additionally, the investigation could be enhanced
by incorporating ocean acidification projections into analyses of coral carbonate budgets.

@@

![Heatmaps indicating reef cluster assignments within each bioregion across the five GCMs.
Green indicates a low cover cluster assignemnt, orange represents a medium cover assignment
and blue represents assignment of a reef to a high coral cover cluster. The vertical width
of reef elements varies depending on the number of reefs within the bioregion (displayed in
the y axis). The x axis represents the different GCM levels (EC-Earth3-Veg, ACCESS-ESM1-5,
ACCESS-CM2, NorESM2-MM and
GFDL-CM4).](../outputs/ADRIA_results/HighResCoralStress/figs/GCM_bioregion_cluster_assignments.png){#sfig-gcm-bioregion-cluster-assignment
width="`{julia} timeseries_width`"}
